\documentclass[10pt,a4paper,onecolumn,article,draft]{memoir}
\counterwithout{section}{chapter}

\input{preamble}

\author{David Miguel Susano Pinto \and Andrew Flaus}
\title{The Human Canonical Core Histone Catalogue}

\begin{document}

  \maketitle

  \begin{abstract}
    Core histone proteins H2A, H2B, H3, and H4 are encoded
    by a large family of genes distributed across the human genome.
    Canonical core histones contribute the majority of proteins to bulk
    chromatin packaging,
    and are encoded in \NumberOfClusters{} clusters in the human genome
    by \TotalCoreCodingGenes{} coding genes comprising
    \HTwoACodingGenes{} for H2A,
    \HTwoBCodingGenes{} for H2B,
    \HThreeCodingGenes{} for H3,
    and \HFourCodingGenes{} for H4,
    along with at least \TotalCorePseudoGenes{} total pseudogenes.
    The canonical core histone genes display coding variation that gives rise to
    \HTwoAUniqueProteins{} H2A, \HTwoBUniqueProteins{} H2B,
    \HThreeUniqueProteins{} H3, and \HFourUniqueProteins{} H4 unique protein isoforms.
    This represents a surprising and seldom recognised variation
    leading to amino acid identity as low as
    \FPround{\result}{\HallMinPID}{0} \SI{\result}{\percent}
    between canonical histones of the same type.
    The gene sequence and protein isoform diversity
    exceeds commonly used subtype designations such as H2A.1 and H3.1,
    and exists in parallel to the well-known specialisation of variant histones.
    To assist understanding of the implications of histone diversity
    we have curated and catalogued the entire human canonical core
    histone gene and protein complement.
    This work has led to multiple corrections to human histone annotations,
    reflecting the flux of the human genome as it is updated and
    enriched in reference databases.
    To avoid obsolescence and maximise accessibility of this information,
    we have applied software build automation tools to
    dynamically generate the canonical histone repertoire
    based on current genome annotations and to insert the information
    into a manuscript format.
    This approach is inspired by the aims of reproducible research
    and can readily be adapted to histones of other organisms and
    to other gene families.
  \end{abstract}

  \input{sections/intro.tex}
  \input{sections/genes.tex}
  \input{sections/transcripts.tex}
  \input{sections/proteins.tex}
  \input{sections/computation.tex}
  \input{sections/conclusion.tex}
  \input{sections/materialsmethods.tex}

  \section{Acknowledgements}
    We are grateful to the Perl community on the \url{irc.freenode.net}
    channels \#perl and \#bioperl,
    especially Altreus, mst, LeoNerd, and pyrimidine.
    Invaluable support was also provided by the \TeX{} stackexchange community,
    in particular David Carlisle.

  \bibliography{references}

  \newpage
  \appendix
  \input{sections/supplementary.tex}

\end{document}
