\documentclass[10pt,a4paper,onecolumn,article,draft]{memoir}
\counterwithout{section}{chapter}
\DoubleSpacing

\usepackage[T1]{fontenc}
\usepackage{palatino}
\usepackage[final]{graphicx}
  \graphicspath{{./figs/}}

%% remove colorlinks when preparing for print
\usepackage[final,hyperindex,hyperfootnotes,bookmarksnumbered,colorlinks]{hyperref}

\usepackage{eqparbox}
  %% keep size of first column of sequence align table on memory to calculate
  %% the size limit for the second column
  \newsavebox{\firstentrybox}
  \newcolumntype{F}{%
    >{\begin{lrbox}{\firstentrybox}}%
      l%
    <{\end{lrbox}%
    \eqmakebox[firstentry][l]{\unhcopy\firstentrybox}}}

\newsubfloat{figure}

\usepackage{color}    % to color background of generated values values
\usepackage{seqsplit}
\usepackage{url}
\usepackage[textsize=footnotesize]{todonotes}
  %% new command for box about missing references
  \newcommand{\addref}[1][]{\todo[color=red!40,size=\tiny]{Add reference: #1}}

\usepackage{natbib}

%% Supplement function from http://bytesizebio.net/2013/03/11/adding-supplementary-tables-and-figures-in-latex/
\newcommand{\beginsupplement}{%
	\setcounter{table}{0}
	\renewcommand{\thetable}{S\arabic{table}}%
	\setcounter{figure}{0}
	\renewcommand{\thefigure}{S\arabic{figure}}%
	}

%% to keep a pair of related table and figures, we set a new float and use it
%%conjunction with the captionof command from capt-of
\usepackage{capt-of}
\newfloat{TableAndFigure}{X}{capname}

\input{results/variables-align_results}
\input{results/variables-cluster_stats}
\input{results/variables-protein_stats}
\input{results/variables-utr}
\input{data/variables-Marzluff-stats}

%% We define a style here which will be used around all the values found
%% by the scripts. The idea is to place them in a grey box so that people
%% reading the paper can really get the idea that the values in the text
%% are automatically generated. The problem is that the variables are not
%% simple numbers or text, and we want to make operations with them here
%% in LaTeX. So we get an handy trick, and redefine FPeval.
%% See http://tex.stackexchange.com/questions/159155 for details
\newcommand{\ScriptValue}[1]{\setlength{\fboxsep}{2pt}\colorbox[gray]{0.8}{#1}}
\let\OldFPeval\FPeval
\def\FPeval{\renewcommand{\ScriptValue}[1]{##1}\OldFPeval}

\author{David Miguel Susano Pinto and Andrew Flaus}
\title{The Human Canonical Core Histone Catalogue}

\begin{document}

  \maketitle

  \begin{abstract}
	Core histone proteins H2A, H2B, H3 and H4 are encoded 
	by a large family of genes distributed across the human genome.
	Canonical core histones contribute the majority of proteins to bulk chromatin packaging, 
	and are encoded by \TotalCodingGenes{} human genes comprising 
	\HTwoACodingGenes{} for H2A,
	\HTwoBCodingGenes{} for H2B,
	\HThreeCodingGenes{} for H3 
	and \HFourCodingGenes{} for H4, 
	along with at least \TotalPseudoGenes{} total pseudogenes.
	The canonical histone genes display coding variation that gives rise to
	\HTwoAUniqueProteins{} H2A, \HTwoBUniqueProteins{} H2B, 
	\HThreeUniqueProteins{} H3 and \HFourUniqueProteins{} unique H4 protein isoforms.
	This represents a surprising and seldom recognised variation
	diverging to as far as \todo{lowest identity}\% amino acid identity. 
	The gene sequence and protein isoform diversity 
	exceeds commonly used subtype designations such as H2A.1 and H3.1
	and exists in parallel to the well-known specialisation of variant histones. 
	To assist understanding of the implications of histone diversity 
	we have curated and catalogued the entire human canonical core histone gene and protein complement.
	This work has led to multiple corrections to human histone annotations, 
	reflecting the flux of the human genome as it is updated and enriched in reference databases. 
	To avoid obsolescence and maximise accessibility of this information, 
	we have developed a novel software framework that dynamically generates the canonical histone repertoire 
	based on current genome annotations and inserts the information into manuscript format. 
	This approach is inspired by the aims of reproducible research 
	and can readily be adapted to histones of other organisms and to other gene families.
  \end{abstract}

  \input{sections/intro.tex}
  \input{sections/genes.tex}
  \input{sections/transcripts.tex}
  \input{sections/proteins.tex}
  \input{sections/computation.tex}
  \input{sections/conclusion.tex}
  \input{sections/materialsmethods.tex}
  \input{sections/supplementary.tex}

  \section{Acknowledgements}
	We are grateful to all the nice people at \#perl and \#bioperl channels on
	\url{irc.freenode.net}, especially Altreus, mst, LeoNerd, and pyrimidine for
	their help with the most technical programming details.

  \bibliographystyle{nar}
  \bibliography{references}

\end{document}
