\documentclass[a4paper,oneside,onecolumn,article,draft]{memoir}
\counterwithout{section}{chapter}

\input{preamble}
\input{preamble-manuscript}

\author{David Miguel Susano Pinto \and Andrew Flaus}
\title{The Human Canonical Core Histone Catalogue}

\begin{document}

  \maketitle

  \begin{abstract}
    Core histone proteins H2A, H2B, H3, and H4 are encoded
    by a large family of genes distributed across the human genome.
    Canonical core histones contribute the majority of proteins to bulk chromatin packaging,
    and are encoded in \NumberOfClusters{} clusters
    by \TotalCoreCodingGenes{} coding genes comprising
    \HTwoACodingGenes{} for H2A,
    \HTwoBCodingGenes{} for H2B,
    \HThreeCodingGenes{} for H3,
    and \HFourCodingGenes{} for H4,
    along with at least \TotalCorePseudoGenes{} total pseudogenes.
    The canonical core histone genes display coding variation that gives rise to
    \HTwoAUniqueProteins{} H2A, \HTwoBUniqueProteins{} H2B,
    \HThreeUniqueProteins{} H3, and \HFourUniqueProteins{} H4 unique protein isoforms.
    Although histone proteins are highly conserved overall,
    these isoforms represent a surprising and seldom recognised variation
    with amino acid identity as low as
    \FPround{\result}{\HallMinPID}{0} \SI{\result}{\percent}
    between canonical histone proteins of the same type.
    The gene sequence and protein isoform diversity
    also exceeds commonly used subtype designations such as H2A.1 and H3.1,
    and exists in parallel with the well-known specialisation of variant histone proteins.
    RNA sequencing of histone transcripts shows evidence for
    differential expression of histone genes
    but the functional significance of this variation has not yet been investigated.
    To assist understanding of the implications of histone gene and protein diversity
    we have curated and catalogued the entire human canonical core
    histone gene and protein complement.
    In order to organise this information in a
    robust, accessible, and accurate form,
    we have applied software build automation tools to
    dynamically generate the canonical core histone repertoire
    based on current genome annotations
    and then to organise the information into a manuscript format.
    Automatically generated values are shown with a light grey background.
    Alongside recognition of the encoded protein diversity,
    this has led to multiple corrections to human histone annotations,
    reflecting the flux of the human genome as it is updated and
    enriched in reference databases.
    This approach is inspired by the aims of reproducible research
    and can be readily adapted to other gene families.
  \end{abstract}

  \input{sections/intro.tex}
  \input{sections/genes.tex}
  \input{sections/transcripts.tex}
  \input{sections/proteins.tex}
  \input{sections/computation.tex}
  \input{sections/conclusion.tex}
  \input{sections/materialsmethods.tex}

  \section{Acknowledgements}
    We are grateful to the Perl community on the \url{irc.freenode.net}
    channels \#perl and \#bioperl,
    especially Altreus, mst, LeoNerd, and pyrimidine.
    Invaluable support was also provided by the \TeX{} stackexchange community,
    in particular David Carlisle.

  \bibliography{references}

  \newpage
  \appendix
  \input{sections/supplementary.tex}

\end{document}
